---
title: "Differential expression - co-houseing mouse mRNA"
author: "Kimberly Olney, PhD"
date: "February 2026"
output:
  pdf_document: default
params:
  project: CH
---

Compare clean, bedding, co-housed by sex

# Set up working enivornment
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
getwd()
```

```{r libraries, message=FALSE, warning=FALSE}
source("/tgen_labs/jfryer/kolney/dirty_mice/dirty_mouse_cohousing/scripts/file_paths_and_colours.R")
projectID <- "CH"
```

# Create a counts matrix
This is the raw counts 
```{r raw_counts}
# path to counts files
count_files <-
  file.path(paste0(
    "/tgen_labs/jfryer/kolney/dirty_mice/dirty_mouse_cohousing/starAligned/",
    metadata$sample, "_STAR.bam", #  "_", metadata$group,"
    typeOfCount
  ))
# add sample counts name to the counts files
names(count_files) <- paste0(metadata$sample)
# Inspect 
table(metadata$sample)

# Create the counts matrix 
count_data <- data.frame(fread(count_files[1]))[c(1,4)] # forth column for RF reads 
# Loop and read the 4th column
for(i in 2:length(count_files)) {
        count_data <- cbind(count_data, data.frame(fread(count_files[i]))[4])
}
# set gene_id as the row name
row.names(count_data) <- count_data$V1
```

# Mapping metrics
```{r mapping_metrics}
star_metrics <- data.frame(count_data[count_data$V1 %like% "N_", ])
star_metrics$V1 <- as.vector(star_metrics$V1)
melt_star_metrics <- reshape::melt(star_metrics, id = c("V1"))
ggplot(melt_star_metrics, aes(x = value, fill = V1)) + 
  geom_histogram(position = "identity", bins = 100) + 
  facet_grid(V1 ~ .)  + theme_bw()
```

clean up
```{r}
remove(melt_star_metrics, star_metrics, i, count_files)
```

# Format counts table
```{r counts_table}
# remove star metric information
count_data <- count_data[!grepl("N_", count_data$V1),]
# set gene_id (i.e. V1) as the row name
row.names(count_data) <- count_data$V1
count_data$V1 <- NULL
# set column names to sample counts file name
colnames(count_data) <- metadata$sample

genes <- read.delim("scripts/ensembl_mouse_genes.txt")
```

# Check inputs
```{r check_inputs}
all.equal(rownames(count_data), genes$gene_id) # gene ids match the count rows
all.equal(colnames(count_data), (metadata$sample)) # counts columns match the sample IDs
```

# Create DGE object
```{r DGE_object}
# group & sex 
metadata$group_sex <- paste0(metadata$group, metadata$sex)
# create object
dge <- DGEList(counts = count_data,
               samples = metadata,
               genes = genes)

# Inspect
table(dge$samples$group)
table(dge$samples$sample)
table(dge$samples$group_sex)

# Create the Female-only object
dge_female <- dge[, dge$samples$sex == "F"]
table(dge_female$samples$group)

# Create the Male-only object
dge_male <- dge[, dge$samples$sex == "M"]
table(dge_male$samples$group)
```

# Set factors and colors 
```{r factor_and_color_set}
dge$samples$group <- factor(dge$samples$group, levels = c("Clean", "Bedding", "CH")) 
condition_colors <- c("skyblue", "orange", "brown")[dge$samples$group]

dge$samples$sex <- factor(dge$samples$sex, levels = c("F", "M"))
sex_colors <- c("purple", "orange")[dge$samples$sex]

# group colors
dge_female$samples$group <- factor(dge_female$samples$group, levels = c("Clean", "Bedding", "CH")) 
condition_colors_female <- c("skyblue", "orange", "brown")[dge_female$samples$group]

dge_male$samples$group <- factor(dge_male$samples$group, levels = c("Clean", "Bedding", "CH")) 
condition_colors_male <- c("skyblue", "orange", "brown")[dge_male$samples$group]
```

# Raw CPM
```{r raw_cpm}
lcpm <- edgeR::cpm(dge$counts, log = TRUE)
lcpm_female <- edgeR::cpm(dge_female$counts, log = TRUE)
lcpm_male <- edgeR::cpm(dge_male$counts, log = TRUE)
```

# Raw MDS
```{r raw_MDS}
# Condition
par(bg = 'white')
plotMDS(
  lcpm_female,
  top = 500,
  labels = dge_female$samples$sample,
  cex = 1,
  dim.plot = c(1, 2),
  plot = TRUE, 
  col = condition_colors_female
)
title(expression('Top 500 Genes - Raw (Log'[2] ~ 'CPM)'))
saveToPDF(paste0("results_", projectID, "/sex_stratified/MDS/MDS_female.pdf"), width = 6, height = 6)

par(bg = 'white')
plotMDS(
  lcpm_male,
  top = 500,
  labels = dge_male$samples$sample,
  cex = 1,
  dim.plot = c(1, 2),
  plot = TRUE, 
  col = condition_colors_male
)
title(expression('Top 500 Genes - Raw (Log'[2] ~ 'CPM)'))
saveToPDF(paste0("results_", projectID, "/sex_stratified/MDS/MDS_male.pdf"), width = 6, height = 6)
```


# Remove mitochondrial genes
```{r MT_genes}
dim(dge) # inspect
removeMT <- dge$genes$Chr != "chrM"  # true when NOT MT
table(removeMT) # inspect
dge <- dge[removeMT,,keep.lib.sizes = FALSE]
dim(dge)

dim(dge_female) # inspect
removeMT <- dge_female$genes$Chr != "chrM"  # true when NOT MT
table(removeMT) # inspect
dge_female <- dge_female[removeMT,,keep.lib.sizes = FALSE]
dim(dge_female)

dim(dge_male) # inspect
removeMT <- dge_male$genes$Chr != "chrM"  # true when NOT MT
table(removeMT) # inspect
dge_male <- dge_male[removeMT,,keep.lib.sizes = FALSE]
dim(dge_male)
```

# Keep only protein coding genes 
```{r MT_genes}
dim(dge)
table(dge$genes$gene_type)
removeMT <- dge$genes$gene_type == "protein_coding"  # true when NOT MT
dge <- dge[removeMT,,keep.lib.sizes = FALSE]
dim(dge)

dim(dge_female)
table(dge_female$genes$gene_type)
removeMT <- dge_female$genes$gene_type == "protein_coding"  # true when NOT MT
dge_female <- dge_female[removeMT,,keep.lib.sizes = FALSE]
dim(dge_female)

dim(dge_male)
table(dge_male$genes$gene_type)
removeMT <- dge_male$genes$gene_type == "protein_coding"  # true when NOT MT
dge_male <- dge_male[removeMT,,keep.lib.sizes = FALSE]
dim(dge_male)
```

# Library sizes
```{r library}
# before filtering
L <- mean(dge$samples$lib.size) * 1e-6
M <- median(dge$samples$lib.size) * 1e-6
c(L, M)

L_female <- mean(dge_female$samples$lib.size) * 1e-6
M_female <- median(dge_female$samples$lib.size) * 1e-6
c(L_female, M_female)

L_male <- mean(dge_male$samples$lib.size) * 1e-6
M_male <- median(dge_male$samples$lib.size) * 1e-6
c(L_male, M_male)
```

# Filtering 
The filterByExpr() function in the edgeR package determines which genes have a great enough count value to keep.  
We will filter by group_sex.  
```{r filter_counts}
# first filter by expression
dim(dge)
keep.expr <-
  filterByExpr(
    dge,
    group = dge$samples$group, # by disease groups
    min.count = M, # min count of 1 CPM 
  )
dge.filtered <- dge[keep.expr, ,keep.lib.sizes = FALSE]
dim(dge.filtered)
table(dge.filtered$genes$gene_type)

dim(dge_female)
keep.expr <-
  filterByExpr(
    dge_female,
    group = dge_female$samples$group, # by disease groups
    min.count = M_female, # min count of 1 CPM 
  )
dge.filtered.female <- dge_female[keep.expr, ,keep.lib.sizes = FALSE]
dim(dge.filtered.female)
table(dge.filtered.female$genes$gene_type)

dim(dge_male)
keep.expr <-
  filterByExpr(
    dge_male,
    group = dge_male$samples$group, # by disease groups
    min.count = M_male, # min count of 1 CPM 
  )
dge.filtered.male <- dge_male[keep.expr, ,keep.lib.sizes = FALSE]
dim(dge.filtered.male)
table(dge.filtered.male$genes$gene_type)
```

# TMM
For estimating relative RNA production levels from RNA-seq data. 
The TMM method estimates scale factors between samples that can be incorporated 
into currently used statistical methods for DE analysis.
```{r TMM}
# Now, method of trimmed mean of M-values (TMM)
# Calculate scaling factors to convert raw library sizes into effective library sizes.
dge.filtered.norm <- calcNormFactors(dge.filtered, method = "TMM")
dge.filtered.norm.female <- calcNormFactors(dge.filtered.female, method = "TMM")
dge.filtered.norm.male <- calcNormFactors(dge.filtered.male, method = "TMM")

# norm factor summary
summary(dge.filtered.norm.female$samples$norm.factors)
normfactors <- (dge.filtered.norm.female$samples$norm.factors)

summary(dge.filtered.norm.male$samples$norm.factors)
normfactors <- (dge.filtered.norm.male$samples$norm.factors)
```

# Gene_id to gene_name
Some of the gene_names are duplicated. Our counts matrix has rownames by gene_id which are not duplicated. Since we have now filtered and normalized we can check to see if duplicated gene_names still exist.  If none do, rename rows to gene_name.  If there are duplicates, handle the error and then rename columns.
```{r gene_name}
# check for duplication
table(duplicated(dge.filtered.norm$genes$gene_name))
table(duplicated(dge.filtered.norm.female$genes$gene_name))
table(duplicated(dge.filtered.norm.male$genes$gene_name))

# replace NA with gene_id 
dge.filtered.norm$genes <- dge.filtered.norm$genes %>% mutate(gene_name = coalesce(gene_name,gene_id))
dge.filtered.norm.female$genes <- dge.filtered.norm.female$genes %>% mutate(gene_name = coalesce(gene_name,gene_id))
dge.filtered.norm.male$genes <- dge.filtered.norm.male$genes %>% mutate(gene_name = coalesce(gene_name,gene_id))

# check for duplication
table(duplicated(dge.filtered.norm$genes$gene_name))
```

# Clean up
```{r cleanup}
rm(count_data, lcpm, keep.expr, normfactors, removeMT, L, M, typeOfCount)
rm(lcpm_female, lcpm_male, L_female, M_female, L_male, M_male, dge.filtered.female, dge.filtered.male, dge_male, dge_female)
```

# Filtered CPM
```{r filtered_lcpm}
cpm <- edgeR::cpm(dge.filtered.norm.female$counts, log = FALSE)
cpm_df <- as.data.frame(cpm)
cpm_df$gene_id <- rownames(cpm_df)
cpm_gene_names <- merge(genes, cpm_df, by = "gene_id")
write.table(cpm_gene_names, paste0("results_", projectID, "/sex_stratified/counts/filtered_cpm_female_gene_names.txt"), sep= "\t", row.names = FALSE, quote=FALSE)
rm(cpm_gene_names, cpm_df, cpm)

cpm <- edgeR::cpm(dge.filtered.norm.male$counts, log = FALSE)
cpm_df <- as.data.frame(cpm)
cpm_df$gene_id <- rownames(cpm_df)
cpm_gene_names <- merge(genes, cpm_df, by = "gene_id")
write.table(cpm_gene_names, paste0("results_", projectID, "/sex_stratified/counts/filtered_cpm_male_gene_names.txt"), sep= "\t", row.names = FALSE, quote=FALSE)
rm(cpm_gene_names, cpm_df, cpm)
```

# Filtered MDS
```{r filtered_MDS}
lcpm_female <- edgeR::cpm(dge.filtered.norm.female$counts, log = FALSE)

# Condition
par(bg = 'white')
plotMDS(
  lcpm_female,
  top = 500,
  labels = dge.filtered.norm.female$samples$sample,
  cex = 1,
  dim.plot = c(1,2),
  plot = TRUE,
  col = condition_colors_female
)
title(expression('Top 500 Genes - Filtered (Log'[2] ~ 'CPM)'))
saveToPDF(paste0("results_", projectID, "/sex_stratified/MDS/MDS_filtered_condition_female.pdf"), width = 6, height = 6)


lcpm_male <- edgeR::cpm(dge.filtered.norm.male$counts, log = FALSE)
# Condition
par(bg = 'white')
plotMDS(
  lcpm_male,
  top = 500,
  labels = dge.filtered.norm.male$samples$sample,
  cex = 1,
  dim.plot = c(1,2),
  plot = TRUE,
  col = condition_colors_male
)
title(expression('Top 500 Genes - Filtered (Log'[2] ~ 'CPM)'))
saveToPDF(paste0("results_", projectID, "/sex_stratified/MDS/MDS_filtered_condition_male.pdf"), width = 6, height = 6)
```


# Voom 
voom transform counts to use for variancePartition  
```{r voom, warning=FALSE}
formula <- (~ 0 + group)
counts_female <- as.data.frame(dge.filtered.norm.female$counts)
info_female <- as.data.frame(dge.filtered.norm.female$samples)
voom_with_weights <-
  variancePartition::voomWithDreamWeights(
    counts = counts_female,
    formula = formula,
    data = info_female,
    BPPARAM = BiocParallel::SnowParam(8),
    plot = TRUE
  )
path <- paste0(paste0("results_", projectID, "/sex_stratified/voom/voom_raw_female"))
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
voomCounts_female <- voom_with_weights$E
rm(voom_with_weights, formula)

formula <- (~ 0 + group)
counts_male <- as.data.frame(dge.filtered.norm.male$counts)
info_male <- as.data.frame(dge.filtered.norm.male$samples)
voom_with_weights <-
  variancePartition::voomWithDreamWeights(
    counts = counts_male,
    formula = formula,
    data = info_male,
    BPPARAM = BiocParallel::SnowParam(8),
    plot = TRUE
  )
path <- paste0(paste0("results_", projectID, "/sex_stratified/voom/voom_raw_male"))
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
voomCounts_male <- voom_with_weights$E
rm(voom_with_weights, formula)
```

# Fit variance 
variancePartition quantifies and interprets multiple sources of biological and technical variation in gene expression experiments. The package a linear mixed model to quantify variation in gene expression attributable to individual, tissue, time point, or technical variables.

```{r varpart}
form_varPart <- ~ (1|group) + (1 | collection_date) + RIN 
# fit model and extract variance percents
rownames(info_female) <- info_female$sample
sample_order_female <- colnames(voomCounts_female)
sorted_info_female <- info_female[match(sample_order_female, info_female$sample), ]
varPart_female <- fitExtractVarPartModel(voomCounts_female, form_varPart, sorted_info_female)

rownames(info_male) <- info_male$sample
sample_order_male <- colnames(voomCounts_male)
sorted_info_male <- info_male[match(sample_order_male, info_male$sample), ]
varPart_male <- fitExtractVarPartModel(voomCounts_male, form_varPart, sorted_info_male)
```

# Plot VarPart
```{r plot_varPart}
plotVarPart(sortCols(varPart_female), label.angle = 80)
saveToPDF(paste0("results_", projectID, "/sex_stratified/variance/variance_partition_violins_female.pdf"), width = 8.5, height = 5)
varPart_female$gene_id <- rownames(varPart_female)
# merge with gene information to get gene names for gene_id
variance_explained_female <- merge(varPart_female, genes, by = "gene_id")
write.table(variance_explained_female, paste0("results_", projectID, "/sex_stratified/variance/variance_explained_female.tsv"),sep = "\t", quote = FALSE, row.names = FALSE)

plotVarPart(sortCols(varPart_male), label.angle = 80)
saveToPDF(paste0("results_", projectID, "/sex_stratified/variance/variance_partition_violins_male.pdf"), width = 8.5, height = 5)
varPart_male$gene_id <- rownames(varPart_male)
# merge with gene information to get gene names for gene_id
variance_explained_male <- merge(varPart_male, genes, by = "gene_id")
write.table(variance_explained_male, paste0("results_", projectID, "/sex_stratified/variance/variance_explained_male.tsv"),sep = "\t", quote = FALSE, row.names = FALSE)
```


# CCA 
Canonical Correlation Analysis
```{r}
form <- ~ group + collection_date + RIN
# Compute Canonical Correlation Analysis (CCA) # between all pairs of variables
# returns absolute correlation value
C = canCorPairs(form, sorted_info_female)
# Plot correlation matrix
plotCorrMatrix( C )
saveToPDF(paste0("results_", projectID, "/sex_stratified/variance/CCA_female.pdf"), width = 5, height = 5)
rm(C)

C = canCorPairs(form, sorted_info_male)
# Plot correlation matrix
plotCorrMatrix( C )
saveToPDF(paste0("results_", projectID, "/sex_stratified/variance/CCA_male.pdf"), width = 5, height = 5)

rm(form, C)
```

# RIN
```{r}
info_female <- dge.filtered.norm.female$samples
p <- info_female %>%
    ggplot(aes(x = group, y = RIN, fill = group)) +
    geom_boxplot() +
    labs(
      title = paste("RIN - female"),
      x = "Experimental Group",
      y = "RIN"
    ) +
    scale_fill_manual(values = c("Clean" = "skyblue", "Bedding" = "orange", "CH" = "brown")) +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, face = "bold"),
     axis.title.x = element_blank() # x-axis title not needed with the facet title
    )
p
saveToPDF(paste0("results_", projectID, "/sex_stratified/variance/RIN_female.pdf"), width = 5, height = 5)


info_male <- dge.filtered.norm.male$samples
p <- info_male %>%
    ggplot(aes(x = group, y = RIN, fill = group)) +
    geom_boxplot() +
    labs(
      title = paste("RIN - male"),
      x = "Experimental Group",
      y = "RIN"
    ) +
    scale_fill_manual(values = c("Clean" = "skyblue", "Bedding" = "orange", "CH" = "brown")) +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, face = "bold"),
     axis.title.x = element_blank() # x-axis title not needed with the facet title
    )
p
saveToPDF(paste0("results_", projectID, "/sex_stratified/variance/RIN_male.pdf"), width = 5, height = 5)
```

# Design matrix
```{r design_matrix}
design_female <- model.matrix( ~ 0 + dge.filtered.norm.female$samples$group + dge.filtered.norm.female$samples$RIN)
colnames(design_female) <- c("Clean", "Bedding", "CH", "RIN")
# Inspect
design_female

design_male <- model.matrix( ~ 0 + dge.filtered.norm.male$samples$group + dge.filtered.norm.male$samples$RIN)
colnames(design_male) <- c("Clean", "Bedding", "CH", "RIN")
# Inspect
design_male
```

# Voom
When the library sizes are quite variable between samples, then the voom approach is theoretically more powerful than limma-trend. 
The voom method estimates the mean-variance relationship of the log-counts.
Generates a precision weight for each observation and enters these into the limma empirical Bayes analysis pipeline.
```{r voom_covariates}
v_female <- voomWithQualityWeights(
  dge.filtered.norm.female,  # our DGEList object
  design_female, # our design matrix
  plot = TRUE # should mean-variance trend & sample-specific weights be displayed?
  )

v_male <- voomWithQualityWeights(
  dge.filtered.norm.male,  # our DGEList object
  design_male, # our design matrix
  plot = TRUE # should mean-variance trend & sample-specific weights be displayed?
  )
```

# Contrast plot
```{r contrasts}
# fits linear model for each gene given a series of arrays
fit_female <- lmFit(v_female, design_female) # voom_cov
coef.fit_female <- fit_female$coefficients

contrasts <- makeContrasts(
  Bedding_vs_Clean = Bedding - Clean,
  CH_vs_Clean = CH - Clean,
  CH_vs_Bedding = CH - Bedding,
  levels = colnames(design_female))
head(contrasts)

# save contrast names
allComparisons <- colnames(contrasts)
allComparisons # check

# run contrast analysis
vfit_female <- contrasts.fit(fit_female, contrasts = contrasts)

# Compute differential expression based on the empirical Bayes moderation of the standard errors towards a common value.
# The logCPM values can then be used in any standard limma pipeline, using the trend=TRUE
veBayesFit_female <- eBayes(vfit_female, trend = TRUE, robust=TRUE)
plotSA(veBayesFit_female, main = "Final Model: Mean-variance Trend")

#male
#rm(design, fit, coef.fit, contrasts, vfit)
fit_male <- lmFit(v_male, design_male) # voom_cov
coef.fit_male <- fit_male$coefficients

contrasts <- makeContrasts(
  Bedding_vs_Clean = Bedding - Clean,
  CH_vs_Clean = CH - Clean,
  CH_vs_Bedding = CH - Bedding,
  levels = colnames(design_male))
head(contrasts)

# save contrast names
allComparisons <- colnames(contrasts)
allComparisons # check

# run contrast analysis
vfit_male <- contrasts.fit(fit_male, contrasts = contrasts)

# Compute differential expression based on the empirical Bayes moderation of the standard errors towards a common value.
# The logCPM values can then be used in any standard limma pipeline, using the trend=TRUE
veBayesFit_male <- eBayes(vfit_male, trend = TRUE, robust=TRUE)
plotSA(veBayesFit_male, main = "Final Model: Mean-variance Trend")
```

# DEGs summary
```{r DGE_summary}
qval <- 0.05
lfc.cutoff <- 0

sumTable_female <- 
  summary(decideTests(
    veBayesFit_female,  # object
    adjust.method = "BH", # by default the method = "separate"
    p.value = qval,
    lfc = lfc.cutoff  # numeric, minimum absolute log2-fold change required
  ))

print(paste0(" q < ", qval, " & absolute log2-fold change > ", lfc.cutoff))
sumTable_female

sumTable_male <- 
  summary(decideTests(
    veBayesFit_male,  # object
    adjust.method = "BH", # by default the method = "separate"
    p.value = qval,
    lfc = lfc.cutoff  # numeric, minimum absolute log2-fold change required
  ))

print(paste0(" q < ", qval, " & absolute log2-fold change > ", lfc.cutoff))
sumTable_male
```

# DEG heatmap 
```{r DEG_heatmap}
sumTable_female <- as.data.frame(sumTable_female)
colnames(sumTable_female) <- c("direction", "comparison", "count")
sumTable_female <- sumTable_female[!sumTable_female$direction == "NotSig",]
sumTable_female <- sumTable_female |> 
  pivot_wider(names_from = comparison, values_from = count)
row_names_to_keep <- sumTable_female$direction
sumTable_female <- sumTable_female[, -which(names(sumTable_female) == "direction")]
sumTable_female <- t(sumTable_female)
colnames(sumTable_female) <- row_names_to_keep
df <- as.data.frame(sumTable_female)

pdf(paste0("results_", projectID, "/sex_stratified/DEGs/DEGs_heatmap_female.pdf"), height = 6, width = 6)
pheatmap::pheatmap(df,
                   main = paste0("q < ", qval, " & |log2FC| > ", lfc.cutoff),
                   treeheight_row = 0,
                   treeheight_col = 0,
                   color = rev(natparks.pals("WindCave")),
                   cluster_rows = FALSE,
                   display_numbers = round(df, digits = 0),
                   fontsize_number = 12,
                   number_color = ifelse(df > 400, "white", "black"),
                   angle_col = "0",
                   breaks = c(seq(0, 500, length.out = length(rev(natparks.pals("WindCave"))))))
dev.off()
rm(df)

sumTable_male <- as.data.frame(sumTable_male)
colnames(sumTable_male) <- c("direction", "comparison", "count")
sumTable_male <- sumTable_male[!sumTable_male$direction == "NotSig",]
sumTable_male <- sumTable_male |> 
  pivot_wider(names_from = comparison, values_from = count)
row_names_to_keep <- sumTable_male$direction
sumTable_male <- sumTable_male[, -which(names(sumTable_male) == "direction")]
sumTable_male <- t(sumTable_male)
colnames(sumTable_male) <- row_names_to_keep
df <- as.data.frame(sumTable_male)

pdf(paste0("results_", projectID, "/sex_stratified/DEGs/DEGs_heatmap_male.pdf"), height = 6, width = 6)
pheatmap::pheatmap(df,
                   main = paste0("q < ", qval, " & |log2FC| > ", lfc.cutoff),
                   treeheight_row = 0,
                   treeheight_col = 0,
                   color = rev(natparks.pals("WindCave")),
                   cluster_rows = FALSE,
                   display_numbers = round(df, digits = 0),
                   fontsize_number = 12,
                   number_color = ifelse(df > 400, "white", "black"),
                   angle_col = "0",
                   breaks = c(seq(0, 500, length.out = length(rev(natparks.pals("WindCave"))))))
dev.off()
```

# Output DEG tables
```{r output_DEG_tables}
genes_relevant_female <- dge.filtered.norm.female$genes

coef <- 1

for (i in allComparisons) {
  vTopTableAll <- topTable(
    veBayesFit_female, 
    coef = coef,  
    n = Inf, 
    p.value = 1,
    lfc = 0, 
    sort.by = "P", 
    genelist = genes_relevant_female, 
    confint = TRUE # column of confidence interval 
    )
    saveRDS(vTopTableAll, file = 
            paste0("rObjects/gene_tables/", i,"_gene_table.rds"))
  path <- paste0("results_", projectID, "/sex_stratified/DEGs/", i, "_q1.00_female.txt", sep = "") 
  write.table(vTopTableAll, path, sep = "\t", row.names = FALSE, quote = FALSE)
  
  # p < 0.05, log2fc > 1
  vTopTableSig <-
    topTable( 
      veBayesFit_female,  
      coef = coef,  
      n = Inf, 
      p.value = qval,
      lfc = lfc.cutoff,
      genelist = genes_relevant_female, 
      confint = TRUE)
  path <- paste0("results_", projectID, "/sex_stratified/DEGs/", i, "_q", qval, "_lfc", lfc.cutoff,"_female.txt", sep = "") 
  write.table(vTopTableSig, path, sep = "\t", row.names = FALSE, quote = FALSE)
  # increment 
  coef <- coef + 1
}
remove(coef, vTopTableAll, vTopTableSig)

genes_relevant_male <- dge.filtered.norm.male$genes

coef <- 1

for (i in allComparisons) {
  vTopTableAll <- topTable(
    veBayesFit_male, 
    coef = coef,  
    n = Inf, 
    p.value = 1,
    lfc = 0, 
    sort.by = "P", 
    genelist = genes_relevant_male, 
    confint = TRUE # column of confidence interval 
    )
    saveRDS(vTopTableAll, file = 
            paste0("rObjects/gene_tables/", i,"_gene_table.rds"))
  path <- paste0("results_", projectID, "/sex_stratified/DEGs/", i, "_q1.00_male.txt", sep = "") 
  write.table(vTopTableAll, path, sep = "\t", row.names = FALSE, quote = FALSE)
  
  # p < 0.05, log2fc > 1
  vTopTableSig <-
    topTable( 
      veBayesFit_male,  
      coef = coef,  
      n = Inf, 
      p.value = qval,
      lfc = lfc.cutoff,
      genelist = genes_relevant_male, 
      confint = TRUE)
  path <- paste0("results_", projectID, "/sex_stratified/DEGs/", i, "_q", qval, "_lfc", lfc.cutoff,"_male.txt", sep = "") 
  write.table(vTopTableSig, path, sep = "\t", row.names = FALSE, quote = FALSE)
  # increment 
  coef <- coef + 1
}
remove(coef, vTopTableAll, vTopTableSig)
```

# PCA
Principal component analysis, or PCA, is a dimensionality reduction method that is often used to reduce the dimensionality of large data sets, by transforming a large set of variables into a smaller one that still contains most of the information in the large set.
```{r PCA}
# Setting the N of genes to use
ntop = length(dge.filtered.norm.female$genes$gene_id)
# Sorting by the coefficient of variance
means <- rowMeans(voomCounts_female)
Pvars <- rowVars(voomCounts_female, useNames = TRUE)
cv2 <- Pvars / means ^ 2
select <- order(cv2, decreasing = TRUE)[seq_len(min(ntop, length(cv2)))]
highly_variable_exp <- ((voomCounts_female)[select,])
dim(highly_variable_exp)

# Running PCA
pca_exp <- prcomp(t(highly_variable_exp), scale = F, center = T)
# scale a logical value indicating whether the variables should be scaled to have unit variance before the analysis takes place.
# a logical value indicating whether the variables should be shifted to be zero centered.

# Dataframe with the first 8 PCs
dim1_10 <- data.frame(pca_exp$x[, 1:8])
# Adding metadata
dim1_10$sample <- rownames(dim1_10)
pcaWithMetadata <- merge(dim1_10, sorted_info_female, by = "sample", all = TRUE)
#pcaWithMetadata$group <- pcaWithMetadata$condition

# Plotting
ggplot(data = pcaWithMetadata, aes(x = PC1, y = PC2, shape = group, color = group)) +
  geom_point(size = 2.5) +
  theme_bw() 
saveToPDF(paste0("results_", projectID, "/sex_stratified/PCA/PCA_dim1&2_female.pdf"), width = 6, height = 5)

# Setting the N of genes to use
ntop = length(dge.filtered.norm.male$genes$gene_id)
# Sorting by the coefficient of variance
means <- rowMeans(voomCounts_male)
Pvars <- rowVars(voomCounts_male, useNames = TRUE)
cv2 <- Pvars / means ^ 2
select <- order(cv2, decreasing = TRUE)[seq_len(min(ntop, length(cv2)))]
highly_variable_exp <- ((voomCounts_male)[select,])
dim(highly_variable_exp)

# Running PCA
pca_exp <- prcomp(t(highly_variable_exp), scale = F, center = T)
# scale a logical value indicating whether the variables should be scaled to have unit variance before the analysis takes place.
# a logical value indicating whether the variables should be shifted to be zero centered.

# Dataframe with the first 8 PCs
dim1_10 <- data.frame(pca_exp$x[, 1:8])
# Adding metadata
dim1_10$sample <- rownames(dim1_10)
pcaWithMetadata <- merge(dim1_10, sorted_info_male, by = "sample", all = TRUE)
#pcaWithMetadata$group <- pcaWithMetadata$condition

# Plotting
ggplot(data = pcaWithMetadata, aes(x = PC1, y = PC2, shape = group, color = group)) +
  geom_point(size = 2.5) +
  theme_bw() 
saveToPDF(paste0("results_", projectID, "/sex_stratified/PCA/PCA_dim1&2_male.pdf"), width = 6, height = 5)
```


# Volcano plots
```{r volcano, warning=FALSE}
negFC <- 0
posFC <- 0
for (i in allComparisons) {
  group1_vs_group2 <-
    read.delim(paste0("results_", projectID, "/sex_stratified/DEGs/", i,"_q1.00_male.txt"))
    i <- gsub("_", " ", i)
    if (!any(group1_vs_group2$adj.P.Val < qval, na.rm = TRUE)) {
    # If there are no significant DEGs, create a volcano plot with all points in gray
    p <- ggplot(data = group1_vs_group2, aes(x = logFC, y = -log10(P.Value))) +
      geom_point(alpha = 0.8, size = 2, color = "gray") +
      theme_bw() +
      theme(legend.position = "none",
            axis.title.x = element_text(size = 10),
            axis.text.x = element_text(size = 10),
            axis.title.y = element_text(size = 10),
            axis.text.y = element_text(size = 10)) +
      geom_vline(xintercept = negFC,
                 colour = "#000000",
                 linetype = "dashed") +
      geom_vline(xintercept = posFC,
                 colour = "#000000",
                 linetype = "dashed") +
      labs(x = expression(log[2](FC)),
           y = expression(-log[10] ~ "(" ~ italic("p") ~ "-value)")) +
      ggtitle(paste0(i, "\nNo significant DEGs found - male"))
    i <- gsub(" ", "_", i)
    pdf(paste0("results_", projectID, "/sex_stratified/volcano/", i, "_male.pdf"), height = 6, width = 6)
    print(p)
    dev.off()
  } else {
  color_values <- vector()
  max <- nrow(group1_vs_group2)
  for (row in 1:max) {
    if (group1_vs_group2$adj.P.Val[row] < qval) {
      if (group1_vs_group2$logFC [row] > posFC) {
        color_values <- c(color_values, 1)
      }
      else if (group1_vs_group2$logFC[row] < negFC) {
        color_values <- c(color_values, 2)
      }
      else {
        color_values <- c(color_values, 3)
      }
    }
    else{
      color_values <- c(color_values, 3)
    }
  }
  group1_vs_group2$color_adjpval_0.05 <- factor(color_values)
  data <- group1_vs_group2
  # plot only if there are DEGs with p_val_adj < 0.05
  num <- subset(data, (adj.P.Val < qval & logFC < negFC)  | (adj.P.Val < qval & logFC > posFC ))
  num <- nrow(num)
  if (num != 0) {
    up <- data[data$color_adjpval_0.05 == 1,]
    up10 <- up[1:10,]
    upFold <- subset(up, logFC > posFC)
    upFold <- upFold[!(upFold$gene_name %in% up10$gene_name),]
    down <- data[data$color_adjpval_0.05 == 2,]
    down10 <- down[1:10,]
    downFold <- subset(down, logFC < negFC)
    downFold <- downFold[!(downFold$gene_name %in% down10$gene_name),]
    if (!1 %in% unique(data$color_adjpval_0.05)) {
      my_colors <- c("blue", "gray")
    } else if (!2 %in% unique(data$color_adjpval_0.05)) {
      my_colors <- c("red", "gray")
    } else if (!1 %in% unique(data$color_adjpval_0.05) &&
               !2 %in% unique(data$color_adjpval_0.05)) {
      my_colors <- c("gray")
    } else {
      my_colors <- c("red", "blue", "gray")
    }
    hadjpval <- (-log10(max(data$P.Value[data$adj.P.Val < qval],
                            na.rm = TRUE)))
    negFC <- c(0)
    posFC <- c(0) 
    p <-
      ggplot(data = data,
             aes(
               x = logFC,
               y = -log10(P.Value),
               color = color_adjpval_0.05
             )) +
      geom_point(alpha = 0.8, size = 1) +
      theme_bw() +
      theme(legend.position = "none") +
      scale_color_manual(values = my_colors) +
      labs(
        title = "",
        x = expression(log[2](FC)),
        y = expression(-log[10] ~ "(" ~ italic("p") ~ "-value)")
      )  +
      geom_hline(yintercept = hadjpval,
                 #  horizontal line
                 colour = "#000000",
                 linetype = "dashed") +
      geom_vline(xintercept = negFC,
                 #  horizontal line
                 colour = "#000000",
                 linetype = "dashed") +
      geom_vline(xintercept = posFC,
                 #  horizontal line
                 colour = "#000000",
                 linetype = "dashed") +
      ggtitle(paste0(i, " male", "\nq < ", qval, " & |log2FC| > ", lfc.cutoff)) + # 
      geom_text_repel(
        data = up10,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = gene_name
        ),
        color = "maroon",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 10)
      ) +
      geom_text_repel(
        data = upFold,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = gene_name
        ),
        color = "maroon",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 10)
      ) +
      geom_text_repel(
        data = down10,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = gene_name
        ),
        color = "navyblue",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 10)
      ) +
      geom_text_repel(
        data = downFold,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = gene_name
        ),
        color = "navyblue",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 10)
      )
    p
    i <- gsub(" ", "_", i)
    # save
    path <-paste0("results_", projectID, "/sex_stratified/volcano/",i)
    pdf(paste0(path, "_male.pdf"),
        height = 6,
        width = 6)
    print(p)
    dev.off()
    
    print(paste("i =", i))
  }
  } 
}
```

```{r cleanup}
# clean up
remove(up, up10, upFold, group1_vs_group2, downFold, down10, data, p, my_colors)
```